---
title: "Marketing Analytics Regression"
author: "Jose Alberto Martinez Morales"
date: "2026-07-31"
output: github_document
subtitle: Regression Modeling of Customer Behavior
---

**Marketing Analytics (DAMO-520-21)**

**Instructor:** Prof. Mohamad Makki

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook applies regression modeling to `marketing_analytics_dataset_5000.csv`
to analyze customer behavior, following the assignment structure:

- **Part A**: Exploratory analysis and three regression models (baseline,
  multiple regression, improved/alternative form)
- **Part B**: Diagnostic testing (linearity, normality, homoscedasticity,
  multicollinearity, influential observations)
- **Part C**: Business implications and limitations (written up separately
  in the final report)

We'll build this up step by step, one chunk at a time.

# Step 0: Load Packages and Data

```{r load-packages}
library(tidyverse)   # dplyr, ggplot2, etc.
library(car)         # VIF, component+residual plots
library(lmtest)      # Breusch-Pagan test
library(GGally)      # correlation matrix / pairs plots
library(readxl)      # for reading .xlsx files
```

```{r load-data}
df <- read_excel("C:/Users/piuze/OneDrive/Documentos/marketing-analytics-regression/data/marketing_analytics_dataset_5000.xlsx")
dim(df)
str(df)
```

# Step 1: Exploratory Analysis

Before selecting a dependent variable and building any regression models, we
first need to understand the dataset itself. This section performs a basic
diagnostic pass over `marketing_analytics_dataset_5000.xlsx` to answer four
questions:

- **Are the data types consistent?** Numeric-looking columns should actually
  be read as numeric, not accidentally stored as text, and vice versa for
  categorical variables.
- **Are there missing values?** If so, how many, and in which columns?
- **Are there duplicate rows?** Repeated customer records would bias any
  model built on top of them.
- **Which variables are continuous, and which are categorical?** This
  distinction matters immediately: continuous variables are candidates for
  direct numeric predictors in a regression model, while categorical
  variables (like `Region` or `Gender`) will need to be encoded before they
  can be used, and binary flags like `PromoUsed` and `Churn` behave more
  like categories than true continuous measurements even though they're
  stored as 0/1 numbers.

Once we understand the shape and quality of the data, we'll visualize the
distribution of every variable, histograms for continuous variables and bar
charts for categorical ones, to get a first look at skewness, spread, and
class imbalance. These visual patterns will directly inform how we justify
our choice of dependent variable and regression type in the next step,
rather than relying on assumptions from theory alone.

**Step 1a: Data Types Check**

```{r dtypes check}
dtypes <- sapply(df, function(x) class(x)[1])
table(dtypes)
```
**Step 1b: Missing Values**

```{r missing values}
colSums(is.na(df))
```
**Step 1c: Duplicates**

```{r duplicates}
cat("Duplicate rows:", sum(duplicated(df)), "\n")
```
**Step 1d: Split Continuous versus Categorical**

```{r variable-types}
id_vars <- c("CustomerID")
categorical_vars <- c("Gender", "Region", "PromoUsed", "Churn")
continuous_vars <- setdiff(names(df), c(id_vars, categorical_vars))

cat("ID variable (excluded from analysis):", paste(id_vars, collapse = ", "), "\n")
cat("Categorical variables:", paste(categorical_vars, collapse = ", "), "\n")
cat("Continuous variables:", paste(continuous_vars, collapse = ", "), "\n")
```

**Step 1e: Histograms for continuous variables**


```{r histograms-plain, fig.width=11, fig.height=8}
df_long <- df %>% select(all_of(continuous_vars)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

ggplot(df_long, aes(x = Value, fill = Variable)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.9) +
  facet_wrap(~Variable, scales = "free") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of Continuous Variables (No Density Curve)",
       x = NULL, y = "Count")
```
**Step 1f: Bar Charts for all Categorical variables**

```{r barcharts, fig.width=9, fig.height=6}
df_long_cat <- df %>% select(all_of(categorical_vars)) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

ggplot(df_long_cat, aes(x = Value, fill = Value)) +
  geom_bar(color = "white", alpha = 0.9) +
  facet_wrap(~Variable, scales = "free_x") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of Categorical Variables", x = NULL, y = "Count")
```
# Step 2: Model Development

*(To fill in together \u2014 Model 1 baseline, Model 2 multiple regression,
Model 3 improved/alternative form.)*

# Part B: Diagnostic Testing

*(To fill in together \u2014 linearity, normality, homoscedasticity,
multicollinearity, influential observations.)*
