---
title: "Marketing Analytics Regression"
author: "Jose Alberto Martinez Morales"
date: "2026-07-31"
output: github_document
subtitle: Regression Modeling of Customer Behavior
---

**Marketing Analytics (DAMO-520-21)**

**Instructor:** Prof. Mohamad Makki

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook applies regression modeling to `marketing_analytics_dataset_5000.csv`
to analyze customer behavior, following the assignment structure:

- **Part A**: Exploratory analysis and three regression models (baseline,
  multiple regression, improved/alternative form)
- **Part B**: Diagnostic testing (linearity, normality, homoscedasticity,
  multicollinearity, influential observations)
- **Part C**: Business implications and limitations (written up separately
  in the final report)

We'll build this up step by step, one chunk at a time.

# Step 0: Load Packages and Data

```{r load-packages}
library(tidyverse)   # dplyr, ggplot2, etc.
library(car)         # VIF, component+residual plots
library(lmtest)      # Breusch-Pagan test
library(GGally)      # correlation matrix / pairs plots
library(readxl)      # for reading .xlsx files
library(patchwork) 
```

```{r load-data}
df <- read_excel("C:/Users/piuze/OneDrive/Documentos/marketing-analytics-regression/data/marketing_analytics_dataset_5000.xlsx")
dim(df)
str(df)
```

# Step 1: Exploratory Analysis

Before selecting a dependent variable and building any regression models, we
first need to understand the dataset itself. This section performs a basic
diagnostic pass over `marketing_analytics_dataset_5000.xlsx` to answer four
questions:

- **Are the data types consistent?** Numeric-looking columns should actually
  be read as numeric, not accidentally stored as text, and vice versa for
  categorical variables.
- **Are there missing values?** If so, how many, and in which columns?
- **Are there duplicate rows?** Repeated customer records would bias any
  model built on top of them.
- **Which variables are continuous, and which are categorical?** This
  distinction matters immediately: continuous variables are candidates for
  direct numeric predictors in a regression model, while categorical
  variables (like `Region` or `Gender`) will need to be encoded before they
  can be used, and binary flags like `PromoUsed` and `Churn` behave more
  like categories than true continuous measurements even though they're
  stored as 0/1 numbers.

Once we understand the shape and quality of the data, we'll visualize the
distribution of every variable, histograms for continuous variables and bar
charts for categorical ones, to get a first look at skewness, spread, and
class imbalance. These visual patterns will directly inform how we justify
our choice of dependent variable and regression type in the next step,
rather than relying on assumptions from theory alone.

**Step 1a: Data Types Check**

```{r dtypes check}
dtypes <- sapply(df, function(x) class(x)[1])
table(dtypes)
```
**Step 1b: Missing Values**

```{r missing values}
colSums(is.na(df))
```
**Step 1c: Duplicates**

```{r duplicates}
cat("Duplicate rows:", sum(duplicated(df)), "\n")
```
**Step 1d: Split Continuous versus Categorical**

```{r variable-types}
id_vars <- c("CustomerID")
categorical_vars <- c("Gender", "Region", "PromoUsed", "Churn")
continuous_vars <- setdiff(names(df), c(id_vars, categorical_vars))

cat("ID variable (excluded from analysis):", paste(id_vars, collapse = ", "), "\n")
cat("Categorical variables:", paste(categorical_vars, collapse = ", "), "\n")
cat("Continuous variables:", paste(continuous_vars, collapse = ", "), "\n")
```

**Step 1e: Histograms for continuous variables**


```{r histograms-plain, fig.width=11, fig.height=8}
df_long <- df %>% select(all_of(continuous_vars)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

ggplot(df_long, aes(x = Value, fill = Variable)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.9) +
  facet_wrap(~Variable, scales = "free") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of Continuous Variables",
       x = NULL, y = "Count")
```
**Step 1f: Bar Charts for all Categorical variables**


```{r barcharts, fig.width=9, fig.height=6}
df_long_cat <- df %>% select(all_of(categorical_vars)) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

ggplot(df_long_cat, aes(x = Value, fill = Value)) +
  geom_bar(color = "white", alpha = 0.9) +
  facet_wrap(~Variable, scales = "free_x") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of Categorical Variables", x = NULL, y = "Count")
```

## Key Insights from Exploratory Analysis

- **Data quality is clean**: no missing values across all 15 columns, no
  duplicate rows, and data types are internally consistent. Of the 4
  categorical variables, only `Gender` and `Region` are stored as character
  columns; `Churn` and `PromoUsed` are categorical in nature (binary
  yes/no flags) but stored as numeric 0/1 values, so a dtypes check alone
  doesn't fully capture the categorical/continuous split. It has to be
  judged by what each variable actually represents.

- **Right-skewed variables**: `Clicks`, `EmailOpens`, `WebsiteVisits`, and
  `TimeOnSite` all show a classic right-skew shape, a peak near low
  values with a long tail extending right. This is typical of count/engagement
  data and suggests these variables are **not normally distributed**,
  a pattern worth citing when justifying a non-linear or transformed model later.

- **Roughly uniform (flat) variables**: `Age`, `AdImpressions`, and
  `PagesViewed` show almost no concentration around a central value, every range is nearly equally likely, unlike a bell curve.

- **Closer to normal**: `Income` and `PurchaseAmount` both show a visibly
  bell-shaped, single-peaked distribution, the strongest candidates for
  variables that behave well under standard linear regression assumptions.

- **Class imbalance in `Churn`**: the majority of customers have `Churn = 0`,
  with far fewer `Churn = 1` cases, important to flag if this variable is
  ever used as a predictor or outcome.

- **Balanced categories elsewhere**: `Gender`, `PromoUsed`, and `Region` are
  all reasonably evenly distributed across their categories, so none of
  these raise a class-imbalance concern.


# Step 2: Business Questions

With the data validated and the basic distributions understood from
Step 1, this section moves from describing the data to actually
*using* it to answer questions a marketing team would care about. Five
questions were selected, each chosen because the straightforward
average or correlation hides a more interesting story underneath, a hidden tension, a decoupled relationship, or an interaction effect
that only shows up once you look at the right combination of
variables.

Each question is paired with one supporting visual and a short
interpretation of what it reveals. We'll build these one at a time,
verifying each visualization actually surfaces the underlying pattern
before moving to the next.

**The five questions:**

1. Are promotions attracting genuine customers, or just bargain
   hunters who churn anyway?
2. Is more time on the website actually worth more money, or is there
   a point of diminishing returns?
3. Which region converts marketing exposure into revenue most
   efficiently?
4. Are the happiest customers also the most valuable, or are
   satisfaction and spending decoupled?
5. Is there a specific age × income combination quietly driving most
   of the revenue?


**Question 1: Are promotions attracting genuine customers, or just bargain hunters who churn anyway ? **


```{r q1-promo-churn, fig.width=10, fig.height=5.5}
summary_promo <- df %>%
  group_by(PromoUsed) %>%
  summarise(avg_purchase = mean(PurchaseAmount), churn_rate = mean(Churn) * 100) %>%
  mutate(PromoUsed = factor(PromoUsed, labels = c("No Promo", "Used Promo")))

p1 <- ggplot(summary_promo, aes(x = PromoUsed, y = avg_purchase, fill = PromoUsed)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = paste0("$", round(avg_purchase, 1))), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("#8FBFE0", "#F4A99B")) +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Average Purchase Amount", x = NULL, y = "Purchase Amount ($)") +
  ylim(0, max(summary_promo$avg_purchase) * 1.2)

p2 <- ggplot(summary_promo, aes(x = PromoUsed, y = churn_rate, fill = PromoUsed)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = paste0(round(churn_rate, 1), "%")), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("#8FBFE0", "#F4A99B")) +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Churn Rate", x = NULL, y = "Churn Rate (%)") +
  ylim(0, max(summary_promo$churn_rate) * 1.3)

p1 + p2 + plot_annotation(title = "Do Promo Users Spend More, But Also Churn More?")
```

**Question 1: Insights**

Promo users spend noticeably more on average ($99.10 vs. $80.40, a 23% lift), while their churn rate is actually slightly lower, not higher (24.2% vs. 24.6%). This directly answers the tension the question was built around: the data does not support the "bargain hunter" concern. If promotions were mainly attracting price-sensitive customers who churn right after redeeming a discount, we'd expect the churn bar to be taller for promo users, instead it's marginally shorter.

This is a genuinely useful marketing finding: **promo usage looks like a clean win**, correlating with both higher spend and (if anything) slightly better retention, rather than trading short-term revenue for long-term churn risk. Worth flagging as a caveat in your write-up though. This is a **correlation, not a causal claim**: customers who were already going to spend more and stay longer might simply be more likely to notice and use promos in the first place (reverse causation / selection effect), rather than the promo itself causing the higher spend.

**Question2: Is more time on the website actually worth more money or is there a point of diminishing returns ?**
  
```{r q2-timeonsite-deciles, fig.width=9, fig.height=6}
decile_summary <- df %>%
  mutate(TimeOnSite_decile = ntile(TimeOnSite, 10)) %>%
  group_by(TimeOnSite_decile) %>%
  summarise(avg_time = mean(TimeOnSite), avg_purchase = mean(PurchaseAmount))

ggplot(decile_summary, aes(x = TimeOnSite_decile, y = avg_purchase)) +
  geom_line(color = "#4A9D5B", linewidth = 1.1) +
  geom_point(size = 3, color = "#4A9D5B") +
  geom_text(aes(label = paste0("$", round(avg_purchase, 0))), vjust = -1.2, size = 3.3) +
  scale_x_continuous(breaks = 1:10) +
  theme_minimal() +
  labs(
    title = "Is There a Point of Diminishing Returns on Time-on-Site?",
    subtitle = "Average Purchase Amount by TimeOnSite Decile (1 = lowest engagement, 10 = highest)",
    x = "TimeOnSite Decile", y = "Average Purchase Amount ($)"
  )
```  


**Question 2: Insights**

The relationship is essentially **flat**, average purchase amount barely moves across `TimeOnSite` deciles, staying in a narrow band between about $88.6 and \$92 (roughly a **4% spread** from lowest to highest decile). That's a meaningfully different pattern than the steadily-climbing curve we saw on the synthetic test data.

This actually answers the business question directly, just not the way we expected: **there's no "point of diminishing returns" to find, because there's no real returns curve here to begin with.** Time spent on the site doesn't appear to meaningfully predict how much a customer spends. The small bump at decile 6 (\$92) is most likely just sampling noise, each decile only contains about 1/10th of your customers, so a single group landing a few dollars higher than its neighbors isn't strong evidence of a real effect, especially when deciles 5, 7, 8, and 9 all sit right back around $89.


**Question 3: Which region converts marketing expose into revenue most efficiently ?**

```{r q3-region-efficiency, fig.width=11, fig.height=5.5}
region_summary <- df %>%
  group_by(Region) %>%
  summarise(
    avg_purchase = mean(PurchaseAmount),
    revenue_per_impression = mean(PurchaseAmount) / mean(AdImpressions)
  ) %>%
  mutate(Region = factor(Region, levels = Region[order(revenue_per_impression)]))

p1 <- ggplot(region_summary, aes(x = Region, y = avg_purchase, fill = Region)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = paste0("$", round(avg_purchase, 0))), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Raw Average Purchase Amount", x = NULL, y = "Purchase Amount ($)") +
  ylim(0, max(region_summary$avg_purchase) * 1.2)

p2 <- ggplot(region_summary, aes(x = Region, y = revenue_per_impression, fill = Region)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = paste0("$", round(revenue_per_impression, 3))), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Revenue per Ad Impression", x = NULL, y = "Revenue / Impression ($)") +
  ylim(0, max(region_summary$revenue_per_impression) * 1.3)

p1 + p2 + plot_annotation(title = "Which Region Is Actually the Most Marketing-Efficient?")
```

**Question 3: Insights**

Here, both panels tell essentially the **same story**, just at different resolutions. Raw purchase amounts are nearly identical across all four regions (\$89–\$90), and revenue-per-impression is similarly tight (\$0.353–\$0.370), but there's a consistent, if modest, ordering in both panels: **North is lowest**, **South is highest**, with East and West sitting in between.

This is the "no hidden story" case we flagged as a possibility: the region generating the most raw revenue (South) is also the most marketing-efficient one, and the region with the lowest raw revenue (North) is also the least efficient. There's no reordering between the two panels, meaning current regional ad allocation isn't obviously *misallocated* relative to where conversion actually happens.

**What this actually tells a marketing team**: the regional differences here are small in absolute terms (a spread of about $0.02 per impression, roughly a 5% gap between North and South), not the dramatic story the "hidden efficiency" framing was hoping to surface. That's a legitimate finding worth stating plainly rather than overselling: **region alone isn't a strong lever for revenue optimization in this dataset**, since all four markets convert ad exposure at roughly comparable rates. If you want a stronger regional story, it might be worth checking whether the *variables driving* purchase amount (like `TimeOnSite` or `Income`) themselves vary meaningfully by region. That could explain why South edges out North even though ad efficiency alone doesn't look dramatically different.

**Question 4: Are the happiest customers also the most valuable or are satisfaction and spending decoupled ?**

```{r q4-satisfaction-churn, fig.width=9.5, fig.height=6.5}
ggplot(df, aes(x = SatisfactionScore, y = PurchaseAmount, color = factor(Churn))) +
  geom_point(alpha = 0.4, size = 1.3) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
  scale_color_manual(values = c("0" = "#5B8FB9", "1" = "#D9534F"),
                      labels = c("Retained", "Churned"), name = "Customer Status") +
  theme_minimal() +
  labs(
    title = "Are Happy Customers Also the Most Valuable — or Are They Decoupled?",
    subtitle = "Purchase Amount vs. Satisfaction, colored by Churn status",
    x = "Satisfaction Score", y = "Purchase Amount ($)"
  )
```

**Question 4: Insights**

This is a clear **"decoupled"** result, both the blue (retained) and red (churned) loess lines are essentially **flat across the entire satisfaction range**, hovering around \$85–\$90 regardless of whether satisfaction is 1 or 5. There's no meaningful upward trend for either group, and the cloud of points is uniformly scattered with no visible funnel or shift as satisfaction increases.

This directly answers the question: **satisfaction and spending are largely decoupled** in this dataset. A customer who rates their satisfaction a 5 isn't spending meaningfully more than one who rates it a 1, and this holds true whether they eventually churned or not. That's actually a fairly important finding, because it pushes back against a natural assumption: it's tempting to treat `SatisfactionScore` as a proxy for customer value, but this chart shows that assumption doesn't hold here.

**Worth flagging carefully, though**: the two lines do diverge very *slightly* at the high-satisfaction end, churned customers trend a touch lower (~$85) while retained customers trend a touch higher (~\$90) around satisfaction = 5. Given how flat and close together both lines are across the rest of the range, I'd treat this small gap as within noise rather than a real signal, unless it's backed up by something more rigorous later (like a regression coefficient with a tight confidence interval). Better to state the finding as "no meaningful relationship between satisfaction and spending" rather than reaching for a subtler story that the data doesn't strongly support.

**Marketing implication**: if satisfaction doesn't predict spend, then investing purely in "keep customers happy" initiatives may not move revenue on its own, retention and upsell strategies might need to target behavioral signals (like `TimeOnSite`, `WebsiteVisits`, or `PromoUsed`, which *did* show real relationships with `PurchaseAmount`) rather than satisfaction scores.

**Question 5 Is there a specific age x income combination quietly driving most of the revenue ?**

```{r q5-age-income-heatmap, fig.width=10, fig.height=7}
heat_data <- df %>%
  mutate(
    age_bracket = cut(Age, breaks = 6, dig.lab = 5),
    income_bracket = cut(Income, breaks = 6, dig.lab = 6)
  ) %>%
  group_by(age_bracket, income_bracket) %>%
  summarise(avg_purchase = mean(PurchaseAmount), n = n(), .groups = "drop")

ggplot(heat_data, aes(x = age_bracket, y = income_bracket, fill = avg_purchase)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0("$", round(avg_purchase, 0))), size = 3.2, color = "black") +
  scale_fill_gradient(low = "#FDEBD0", high = "#B9481F", name = "Avg Purchase ($)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  labs(
    title = "Is There a Hidden Age x Income Interaction Driving Revenue?",
    subtitle = "Average Purchase Amount by Age Bracket x Income Bracket",
    x = "Age Bracket", y = "Income Bracket"
  )
```


**Question 5: Insights** 

This is a clean, clear pattern: **strong vertical gradient, flat horizontal gradient**. Moving up any single column (increasing income) takes you from ~\$68 to ~\$130+, a massive, consistent climb. Moving across any single row (changing age, holding income roughly constant) barely moves the number at all, most rows vary by only $1–3 across all six age brackets.

**This answers Question 5 directly, but not with a "yes" to the hidden-interaction hypothesis**: there's no special age × income combination secretly driving revenue. Instead, the heatmap reveals something simpler and just as useful, **income is doing essentially all the work, and age is doing almost none**. This lines up with what we already suspected from Step 1's insight that `Age` showed a flat, uniform distribution with near-zero correlation to `PurchaseAmount` on its own; this chart confirms that flatness holds true even *within* every income level, not just in aggregate.

**One cell worth a second look, but with a caveat**: the top-right cell (\$137, for ages 55.3–64.7 in the highest income bracket) sits noticeably above its row-neighbors, which are otherwise tightly clustered around \$124. That could hint at a genuine "older, high-income" premium segment, but the top income bracket almost certainly has the smallest sample size of any cell in this grid (high earners are rarer), so a single elevated cell there could easily be noise rather than a real interaction. 

**Marketing implication**: since income is clearly the dominant driver and age contributes essentially nothing, segmentation and targeting strategies should prioritize income tier over age group, age-based marketing personas may not be adding much predictive value here.



# Part B: Diagnostic Testing

*(To fill in together \u2014 linearity, normality, homoscedasticity,
multicollinearity, influential observations.)*
