---
title: "Marketing Analytics Regression"
author: "Jose Alberto Martinez Morales"
date: "2026-07-31"
output: github_document
subtitle: Regression Modeling of Customer Behavior
---

**Marketing Analytics (DAMO-520-21)**

**Instructor:** Prof. Mohamad Makki

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook applies regression modeling to `marketing_analytics_dataset_5000.csv`
to analyze customer behavior, following the assignment structure:

- **Part A**: Exploratory analysis and three regression models (baseline,
  multiple regression, improved/alternative form)
- **Part B**: Diagnostic testing (linearity, normality, homoscedasticity,
  multicollinearity, influential observations)
- **Part C**: Business implications and limitations (written up separately
  in the final report)

We'll build this up step by step, one chunk at a time.

# Step 0: Load Packages and Data

```{r load-packages}
library(tidyverse)   # dplyr, ggplot2, etc.
library(car)         # VIF, component+residual plots
library(lmtest)      # Breusch-Pagan test
library(GGally)      # correlation matrix / pairs plots
library(readxl)      # for reading .xlsx files
library(patchwork) 
```

```{r load-data}
df <- read_excel("C:/Users/piuze/OneDrive/Documentos/marketing-analytics-regression/data/marketing_analytics_dataset_5000.xlsx")
dim(df)
str(df)
```

# Step 1: Exploratory Analysis

Before selecting a dependent variable and building any regression models, we
first need to understand the dataset itself. This section performs a basic
diagnostic pass over `marketing_analytics_dataset_5000.xlsx` to answer four
questions:

- **Are the data types consistent?** Numeric-looking columns should actually
  be read as numeric, not accidentally stored as text, and vice versa for
  categorical variables.
- **Are there missing values?** If so, how many, and in which columns?
- **Are there duplicate rows?** Repeated customer records would bias any
  model built on top of them.
- **Which variables are continuous, and which are categorical?** This
  distinction matters immediately: continuous variables are candidates for
  direct numeric predictors in a regression model, while categorical
  variables (like `Region` or `Gender`) will need to be encoded before they
  can be used, and binary flags like `PromoUsed` and `Churn` behave more
  like categories than true continuous measurements even though they're
  stored as 0/1 numbers.

Once we understand the shape and quality of the data, we'll visualize the
distribution of every variable, histograms for continuous variables and bar
charts for categorical ones, to get a first look at skewness, spread, and
class imbalance. These visual patterns will directly inform how we justify
our choice of dependent variable and regression type in the next step,
rather than relying on assumptions from theory alone.

**Step 1a: Data Types Check**

```{r dtypes check}
dtypes <- sapply(df, function(x) class(x)[1])
table(dtypes)
```
**Step 1b: Missing Values**

```{r missing values}
colSums(is.na(df))
```
**Step 1c: Duplicates**

```{r duplicates}
cat("Duplicate rows:", sum(duplicated(df)), "\n")
```
**Step 1d: Split Continuous versus Categorical**

```{r variable-types}
id_vars <- c("CustomerID")
categorical_vars <- c("Gender", "Region", "PromoUsed", "Churn")
continuous_vars <- setdiff(names(df), c(id_vars, categorical_vars))

cat("ID variable (excluded from analysis):", paste(id_vars, collapse = ", "), "\n")
cat("Categorical variables:", paste(categorical_vars, collapse = ", "), "\n")
cat("Continuous variables:", paste(continuous_vars, collapse = ", "), "\n")
```

**Step 1e: Histograms for continuous variables**


```{r histograms-plain, fig.width=11, fig.height=8}
df_long <- df %>% select(all_of(continuous_vars)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

ggplot(df_long, aes(x = Value, fill = Variable)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.9) +
  facet_wrap(~Variable, scales = "free") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of Continuous Variables",
       x = NULL, y = "Count")
```
**Step 1f: Bar Charts for all Categorical variables**


```{r barcharts, fig.width=9, fig.height=6}
df_long_cat <- df %>% select(all_of(categorical_vars)) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

ggplot(df_long_cat, aes(x = Value, fill = Value)) +
  geom_bar(color = "white", alpha = 0.9) +
  facet_wrap(~Variable, scales = "free_x") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of Categorical Variables", x = NULL, y = "Count")
```

## Key Insights from Exploratory Analysis

- **Data quality is clean**: no missing values across all 15 columns, no
  duplicate rows, and data types are internally consistent. Of the 4
  categorical variables, only `Gender` and `Region` are stored as character
  columns; `Churn` and `PromoUsed` are categorical in nature (binary
  yes/no flags) but stored as numeric 0/1 values, so a dtypes check alone
  doesn't fully capture the categorical/continuous split. It has to be
  judged by what each variable actually represents.

- **Right-skewed variables**: `Clicks`, `EmailOpens`, `WebsiteVisits`, and
  `TimeOnSite` all show a classic right-skew shape, a peak near low
  values with a long tail extending right. This is typical of count/engagement
  data and suggests these variables are **not normally distributed**,
  a pattern worth citing when justifying a non-linear or transformed model later.

- **Roughly uniform (flat) variables**: `Age`, `AdImpressions`, and
  `PagesViewed` show almost no concentration around a central value, every range is nearly equally likely, unlike a bell curve.

- **Closer to normal**: `Income` and `PurchaseAmount` both show a visibly
  bell-shaped, single-peaked distribution, the strongest candidates for
  variables that behave well under standard linear regression assumptions.

- **Class imbalance in `Churn`**: the majority of customers have `Churn = 0`,
  with far fewer `Churn = 1` cases, important to flag if this variable is
  ever used as a predictor or outcome.

- **Balanced categories elsewhere**: `Gender`, `PromoUsed`, and `Region` are
  all reasonably evenly distributed across their categories, so none of
  these raise a class-imbalance concern.


# Step 2: Business Questions

With the data validated and the basic distributions understood from
Step 1, this section moves from describing the data to actually
*using* it to answer questions a marketing team would care about. Five
questions were selected, each chosen because the straightforward
average or correlation hides a more interesting story underneath, a hidden tension, a decoupled relationship, or an interaction effect
that only shows up once you look at the right combination of
variables.

Each question is paired with one supporting visual and a short
interpretation of what it reveals. We'll build these one at a time,
verifying each visualization actually surfaces the underlying pattern
before moving to the next.

**The five questions:**

1. Are promotions attracting genuine customers, or just bargain
   hunters who churn anyway?
2. Is more time on the website actually worth more money, or is there
   a point of diminishing returns?
3. Which region converts marketing exposure into revenue most
   efficiently?
4. Are the happiest customers also the most valuable, or are
   satisfaction and spending decoupled?
5. Is there a specific age × income combination quietly driving most
   of the revenue?


**Question 1: Are promotions attracting genuine customers, or just bargain hunters who churn anyway ? **


```{r q1-promo-churn, fig.width=10, fig.height=5.5}
summary_promo <- df %>%
  group_by(PromoUsed) %>%
  summarise(avg_purchase = mean(PurchaseAmount), churn_rate = mean(Churn) * 100) %>%
  mutate(PromoUsed = factor(PromoUsed, labels = c("No Promo", "Used Promo")))

p1 <- ggplot(summary_promo, aes(x = PromoUsed, y = avg_purchase, fill = PromoUsed)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = paste0("$", round(avg_purchase, 1))), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("#8FBFE0", "#F4A99B")) +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Average Purchase Amount", x = NULL, y = "Purchase Amount ($)") +
  ylim(0, max(summary_promo$avg_purchase) * 1.2)

p2 <- ggplot(summary_promo, aes(x = PromoUsed, y = churn_rate, fill = PromoUsed)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = paste0(round(churn_rate, 1), "%")), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("#8FBFE0", "#F4A99B")) +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Churn Rate", x = NULL, y = "Churn Rate (%)") +
  ylim(0, max(summary_promo$churn_rate) * 1.3)

p1 + p2 + plot_annotation(title = "Do Promo Users Spend More, But Also Churn More?")
```

**Question 1: Insights**

Promo users spend noticeably more on average ($99.10 vs. $80.40, a 23% lift), while their churn rate is actually slightly lower, not higher (24.2% vs. 24.6%). This directly answers the tension the question was built around: the data does not support the "bargain hunter" concern. If promotions were mainly attracting price-sensitive customers who churn right after redeeming a discount, we'd expect the churn bar to be taller for promo users, instead it's marginally shorter.

This is a genuinely useful marketing finding: **promo usage looks like a clean win**, correlating with both higher spend and (if anything) slightly better retention, rather than trading short-term revenue for long-term churn risk. Worth flagging as a caveat in your write-up though. This is a **correlation, not a causal claim**: customers who were already going to spend more and stay longer might simply be more likely to notice and use promos in the first place (reverse causation / selection effect), rather than the promo itself causing the higher spend.

**Question2: Is more time on the website actually worth more money or is there a point of diminishing returns ?**
  
```{r q2-timeonsite-deciles, fig.width=9, fig.height=6}
decile_summary <- df %>%
  mutate(TimeOnSite_decile = ntile(TimeOnSite, 10)) %>%
  group_by(TimeOnSite_decile) %>%
  summarise(avg_time = mean(TimeOnSite), avg_purchase = mean(PurchaseAmount))

ggplot(decile_summary, aes(x = TimeOnSite_decile, y = avg_purchase)) +
  geom_line(color = "#4A9D5B", linewidth = 1.1) +
  geom_point(size = 3, color = "#4A9D5B") +
  geom_text(aes(label = paste0("$", round(avg_purchase, 0))), vjust = -1.2, size = 3.3) +
  scale_x_continuous(breaks = 1:10) +
  theme_minimal() +
  labs(
    title = "Is There a Point of Diminishing Returns on Time-on-Site?",
    subtitle = "Average Purchase Amount by TimeOnSite Decile (1 = lowest engagement, 10 = highest)",
    x = "TimeOnSite Decile", y = "Average Purchase Amount ($)"
  )
```  


**Question 2: Insights**

The relationship is essentially **flat**, average purchase amount barely moves across `TimeOnSite` deciles, staying in a narrow band between about $88.6 and \$92 (roughly a **4% spread** from lowest to highest decile). That's a meaningfully different pattern than the steadily-climbing curve we saw on the synthetic test data.

This actually answers the business question directly, just not the way we expected: **there's no "point of diminishing returns" to find, because there's no real returns curve here to begin with.** Time spent on the site doesn't appear to meaningfully predict how much a customer spends. The small bump at decile 6 (\$92) is most likely just sampling noise, each decile only contains about 1/10th of your customers, so a single group landing a few dollars higher than its neighbors isn't strong evidence of a real effect, especially when deciles 5, 7, 8, and 9 all sit right back around $89.


**Question 3: Which region converts marketing expose into revenue most efficiently ?**

```{r q3-region-efficiency, fig.width=11, fig.height=5.5}
region_summary <- df %>%
  group_by(Region) %>%
  summarise(
    avg_purchase = mean(PurchaseAmount),
    revenue_per_impression = mean(PurchaseAmount) / mean(AdImpressions)
  ) %>%
  mutate(Region = factor(Region, levels = Region[order(revenue_per_impression)]))

p1 <- ggplot(region_summary, aes(x = Region, y = avg_purchase, fill = Region)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = paste0("$", round(avg_purchase, 0))), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Raw Average Purchase Amount", x = NULL, y = "Purchase Amount ($)") +
  ylim(0, max(region_summary$avg_purchase) * 1.2)

p2 <- ggplot(region_summary, aes(x = Region, y = revenue_per_impression, fill = Region)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = paste0("$", round(revenue_per_impression, 3))), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Revenue per Ad Impression", x = NULL, y = "Revenue / Impression ($)") +
  ylim(0, max(region_summary$revenue_per_impression) * 1.3)

p1 + p2 + plot_annotation(title = "Which Region Is Actually the Most Marketing-Efficient?")
```

**Question 3: Insights**

Here, both panels tell essentially the **same story**, just at different resolutions. Raw purchase amounts are nearly identical across all four regions (\$89–\$90), and revenue-per-impression is similarly tight (\$0.353–\$0.370), but there's a consistent, if modest, ordering in both panels: **North is lowest**, **South is highest**, with East and West sitting in between.

This is the "no hidden story" case we flagged as a possibility: the region generating the most raw revenue (South) is also the most marketing-efficient one, and the region with the lowest raw revenue (North) is also the least efficient. There's no reordering between the two panels, meaning current regional ad allocation isn't obviously *misallocated* relative to where conversion actually happens.

**What this actually tells a marketing team**: the regional differences here are small in absolute terms (a spread of about $0.02 per impression, roughly a 5% gap between North and South), not the dramatic story the "hidden efficiency" framing was hoping to surface. That's a legitimate finding worth stating plainly rather than overselling: **region alone isn't a strong lever for revenue optimization in this dataset**, since all four markets convert ad exposure at roughly comparable rates. If you want a stronger regional story, it might be worth checking whether the *variables driving* purchase amount (like `TimeOnSite` or `Income`) themselves vary meaningfully by region. That could explain why South edges out North even though ad efficiency alone doesn't look dramatically different.

**Question 4: Are the happiest customers also the most valuable or are satisfaction and spending decoupled ?**

```{r q4-satisfaction-churn, fig.width=9.5, fig.height=6.5}
ggplot(df, aes(x = SatisfactionScore, y = PurchaseAmount, color = factor(Churn))) +
  geom_point(alpha = 0.4, size = 1.3) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
  scale_color_manual(values = c("0" = "#5B8FB9", "1" = "#D9534F"),
                      labels = c("Retained", "Churned"), name = "Customer Status") +
  theme_minimal() +
  labs(
    title = "Are Happy Customers Also the Most Valuable — or Are They Decoupled?",
    subtitle = "Purchase Amount vs. Satisfaction, colored by Churn status",
    x = "Satisfaction Score", y = "Purchase Amount ($)"
  )
```

**Question 4: Insights**

This is a clear **"decoupled"** result, both the blue (retained) and red (churned) loess lines are essentially **flat across the entire satisfaction range**, hovering around \$85–\$90 regardless of whether satisfaction is 1 or 5. There's no meaningful upward trend for either group, and the cloud of points is uniformly scattered with no visible funnel or shift as satisfaction increases.

This directly answers the question: **satisfaction and spending are largely decoupled** in this dataset. A customer who rates their satisfaction a 5 isn't spending meaningfully more than one who rates it a 1, and this holds true whether they eventually churned or not. That's actually a fairly important finding, because it pushes back against a natural assumption: it's tempting to treat `SatisfactionScore` as a proxy for customer value, but this chart shows that assumption doesn't hold here.

**Worth flagging carefully, though**: the two lines do diverge very *slightly* at the high-satisfaction end, churned customers trend a touch lower (~$85) while retained customers trend a touch higher (~\$90) around satisfaction = 5. Given how flat and close together both lines are across the rest of the range, I'd treat this small gap as within noise rather than a real signal, unless it's backed up by something more rigorous later (like a regression coefficient with a tight confidence interval). Better to state the finding as "no meaningful relationship between satisfaction and spending" rather than reaching for a subtler story that the data doesn't strongly support.

**Marketing implication**: if satisfaction doesn't predict spend, then investing purely in "keep customers happy" initiatives may not move revenue on its own, retention and upsell strategies might need to target behavioral signals (like `TimeOnSite`, `WebsiteVisits`, or `PromoUsed`, which *did* show real relationships with `PurchaseAmount`) rather than satisfaction scores.

**Question 5 Is there a specific age x income combination quietly driving most of the revenue ?**

```{r q5-age-income-heatmap, fig.width=10, fig.height=7}
heat_data <- df %>%
  mutate(
    age_bracket = cut(Age, breaks = 6, dig.lab = 5),
    income_bracket = cut(Income, breaks = 6, dig.lab = 6)
  ) %>%
  group_by(age_bracket, income_bracket) %>%
  summarise(avg_purchase = mean(PurchaseAmount), n = n(), .groups = "drop")

ggplot(heat_data, aes(x = age_bracket, y = income_bracket, fill = avg_purchase)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0("$", round(avg_purchase, 0))), size = 3.2, color = "black") +
  scale_fill_gradient(low = "#FDEBD0", high = "#B9481F", name = "Avg Purchase ($)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  labs(
    title = "Is There a Hidden Age x Income Interaction Driving Revenue?",
    subtitle = "Average Purchase Amount by Age Bracket x Income Bracket",
    x = "Age Bracket", y = "Income Bracket"
  )
```


**Question 5: Insights** 

This is a clean, clear pattern: **strong vertical gradient, flat horizontal gradient**. Moving up any single column (increasing income) takes you from ~\$68 to ~\$130+, a massive, consistent climb. Moving across any single row (changing age, holding income roughly constant) barely moves the number at all, most rows vary by only $1–3 across all six age brackets.

**This answers Question 5 directly, but not with a "yes" to the hidden-interaction hypothesis**: there's no special age × income combination secretly driving revenue. Instead, the heatmap reveals something simpler and just as useful, **income is doing essentially all the work, and age is doing almost none**. This lines up with what we already suspected from Step 1's insight that `Age` showed a flat, uniform distribution with near-zero correlation to `PurchaseAmount` on its own; this chart confirms that flatness holds true even *within* every income level, not just in aggregate.

**One cell worth a second look, but with a caveat**: the top-right cell (\$137, for ages 55.3–64.7 in the highest income bracket) sits noticeably above its row-neighbors, which are otherwise tightly clustered around \$124. That could hint at a genuine "older, high-income" premium segment, but the top income bracket almost certainly has the smallest sample size of any cell in this grid (high earners are rarer), so a single elevated cell there could easily be noise rather than a real interaction. 

**Marketing implication**: since income is clearly the dominant driver and age contributes essentially nothing, segmentation and targeting strategies should prioritize income tier over age group, age-based marketing personas may not be adding much predictive value here.


# Step 3: Outlier Detection

Before moving into regression modeling, it's worth taking a closer,
more formal look at outliers, since they matter more for a linear
regression model than almost any other analysis step so far. Ordinary
least squares regression fits a line by *minimizing squared residuals*,
which means a single extreme value can pull the entire fitted line
toward it far more aggressively than it would under a method that
doesn't square its errors. In practice, this shows up in several ways:

- **Inflated or deflated coefficients**, a handful of extreme points
  can shift a slope enough to misrepresent the relationship for the
  other 99% of the data
- **Violated assumptions**, outliers are a common cause of
  heteroscedasticity (non-constant error variance) and non-normal
  residuals, both of which we'll test formally in Part B
- **Misleading R-squared**, a model can appear to fit well overall
  while badly misfitting the bulk of ordinary observations, simply
  because a few extreme points anchor the fit
- **Reduced generalizability**, a model overly influenced by rare
  extreme cases may perform poorly on new, typical customers, which
  is the opposite of what a business wants from a predictive model

Because of this, choosing the *right way* to detect outliers matters
almost as much as detecting them at all. Outlier detection methods
generally fall into two families, and which one is appropriate depends
on the shape of each variable's distribution:

- **Z-score method**, flags a value as an outlier if it falls more
  than a fixed number of standard deviations (typically ±3) from the
  mean. This method implicitly assumes the variable is **approximately
  normally distributed**, since it relies on the mean and standard
  deviation as meaningful reference points. Applying it to a skewed
  variable can badly under, or over-flag outliers, since the mean
  itself is pulled by the skew.
- **IQR (interquartile range) method**, flags a value as an outlier
  if it falls more than 1.5 times the interquartile range below the
  first quartile or above the third. This method is **distribution-
  agnostic** and far more robust to skewed data, since it's built on
  quartiles rather than the mean.

This is exactly why **skewness** is checked first in this section,
essentially a numeric extension of the visual distribution shapes we
already saw in Step 1's histograms (recall the clear right-skew in
`TimeOnSite`, `Clicks`, `WebsiteVisits`, and `EmailOpens`, versus the
closer-to-normal shape of `Income` and `PurchaseAmount`). A skewness
value near 0 is a strong indicator a variable is close to normally
distributed, making the z-score method appropriate; a skewness value
meaningfully above or below 0 signals the distribution is lopsided,
in which case the **IQR method** is the safer, more defensible choice.

The workflow for this section, then, is:

1. Calculate skewness for every continuous variable
2. For variables close to normal (low skewness), apply the **z-score**
   method for outlier detection
3. For variables that are meaningfully skewed, apply the **IQR**
   method instead
4. Report outlier counts per variable, and flag (without necessarily
   removing) any variable with a high proportion of outliers before
   moving into regression modeling


**Skewness Calculation plus Conditional Method Selection**

```{r step3-skewness-detection}
library(e1071)

skew_vals <- sapply(df[, continuous_vars], skewness)

count_outliers_zscore <- function(x) {
  z <- (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
  sum(abs(z) > 3, na.rm = TRUE)
}

count_outliers_iqr <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE); q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  sum(x < (q1 - 1.5 * iqr) | x > (q3 + 1.5 * iqr), na.rm = TRUE)
}

skew_threshold <- 0.5  # |skewness| below this is treated as "near-normal"

outlier_results <- data.frame(Variable = continuous_vars, Skewness = round(skew_vals, 3)) %>%
  mutate(
    Method = ifelse(abs(Skewness) < skew_threshold, "Z-score (near-normal)", "IQR (skewed)"),
    Outliers = mapply(function(v, m) {
      x <- df[[v]]
      if (m == "Z-score (near-normal)") count_outliers_zscore(x) else count_outliers_iqr(x)
    }, Variable, Method),
    Pct = round(Outliers / nrow(df) * 100, 2)
  ) %>%
  arrange(desc(Outliers))

print(outlier_results, row.names = FALSE)
```
**Visualization: Skewness plus Resulting Outlier Counts, side by sid**

```{r step3-outlier-viz, fig.width=13, fig.height=6.5}
library(patchwork)

results_sorted1 <- outlier_results %>% mutate(Variable = factor(Variable, levels = Variable[order(Skewness)]))

p1 <- ggplot(results_sorted1, aes(x = Skewness, y = Variable, color = Method)) +
  geom_segment(aes(x = 0, xend = Skewness, y = Variable, yend = Variable), linewidth = 1.1, alpha = 0.7) +
  geom_point(size = 4) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60", linewidth = 0.4) +
  scale_color_manual(values = c("Z-score (near-normal)" = "#A7D3A0", "IQR (skewed)" = "#F4B8B0")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(title = "Skewness by Variable", subtitle = "Dashed lines mark the |0.5| near-normal threshold",
       x = "Skewness", y = NULL, color = "Method Triggered")

results_sorted2 <- outlier_results %>% mutate(Variable = factor(Variable, levels = Variable[order(Outliers)]))

p2 <- ggplot(results_sorted2, aes(x = Outliers, y = Variable, fill = Method)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = Outliers), hjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("Z-score (near-normal)" = "#A7D3A0", "IQR (skewed)" = "#F4B8B0")) +
  theme_minimal() +
  theme(legend.position = "none") +
  xlim(0, max(results_sorted2$Outliers) * 1.25) +
  labs(title = "Outliers Detected", subtitle = "Using the method matched to each distribution shape",
       x = "Outlier Count", y = NULL)

p1 + p2 + plot_annotation(
  title = "Step 3: Outlier Detection. Kewness-Guided Method Selection",
  theme = theme(plot.title = element_text(size = 15, face = "bold"))
)
```

## Step 3: Insights from Outlier Detection

**Almost every variable in this dataset is close to symmetric.** Only
one variable, `Clicks` (skewness = 0.577), crossed the ±0.5 threshold
into "skewed" territory and triggered the IQR method. Every other
variable, including `TimeOnSite` (skewness = 0.117), which showed a
visibly right-skewed histogram back in Step 1, turned out to be much
closer to normal than expected once measured numerically. This is a
useful reminder that a histogram's *visual* shape can look more skewed
than the actual skewness statistic supports, especially with a large
sample size, which is exactly why we calculate skewness formally
instead of relying on eyeballing the shape alone.

**Overall outlier prevalence is very low across the board.** The
highest outlier count belongs to `Clicks` at just 53 rows (1.06% of
the dataset), and five variables (`Age`, `PagesViewed`, `AdImpressions`,
`EmailOpens`, `SatisfactionScore`) show **zero** outliers under either
method. This is a genuinely reassuring result heading into regression
modeling: with no variable showing more than roughly 1% extreme values,
outliers are unlikely to be a dominant force distorting model
coefficients later in Part A.

**A caution on `PurchaseAmount` specifically.** Even though its outlier
count is small (10 rows, 0.20%), this is the **dependent variable**,
any outlier here isn't just a data quirk to note in passing, it directly
shapes the regression line itself, since OLS minimizes *squared*
residuals and a handful of extreme purchase values can pull the fitted
line disproportionately toward them. Before deciding whether to keep,
transform, or investigate these 10 rows, it's worth looking at them
individually: are they legitimately large purchases from real
high-value customers, or do they look like data entry errors (e.g., an
extra zero)? If they're genuine, they likely represent real revenue
behavior worth preserving in the model rather than treating as noise.

**A similar caution applies to `Clicks` and `WebsiteVisits`.** These
outliers may not be errors at all; they could represent a small
segment of unusually engaged users, the "power users" who click and
browse far more than a typical customer. Removing them purely because
they're statistical outliers risks discarding exactly the behavioral
signal a marketing team might care about most (e.g., identifying
what drives high engagement). The safer approach, consistent with the
reasoning from Step 3's introduction, is to **flag these counts and
inspect the affected rows, not to delete them automatically**, a
decision to exclude any of them should be justified by evidence of a
data error, not just by the statistical rule that flagged them.

**Net takeaway for Part A:** with such low outlier prevalence overall,
this dataset does not appear to have a systemic outlier problem that
would force an early transformation or exclusion strategy. Any
decision to treat specific rows differently should be targeted and
well-justified (especially for `PurchaseAmount`), rather than applying
a blanket removal rule across the dataset.


# Step 4: Baseline Model (1 Predictor)

With the data cleaned, explored, and outlier behavior understood, this
section builds the first regression model: a **baseline model using a
single predictor**. The purpose of a baseline model isn't to be the
"best" model, it's to establish a simple, interpretable reference point
that every later, more complex model (Model 2's multiple regression,
Model 3's improved form) must actually outperform to justify its added
complexity.

**Why `Income` was chosen as the single predictor, based on evidence
rather than assumption:**

The assignment explicitly requires justifying regression choices
"based on your actual data, not theory alone," and two pieces of
evidence from the business-question EDA in Step 2 point clearly to
`Income`:

- The **age × income heatmap** (Question 5) showed a strong, consistent
  gradient in `PurchaseAmount` moving across income tiers, while the
  same chart showed almost no movement across age brackets at any
  fixed income level. Income, not age, was doing essentially all of
  the explanatory work.
- The **time-on-site decile analysis** (Question 2), a variable that
  looked visually promising in Step 1's histograms, turned out to be
  essentially flat against `PurchaseAmount` once tested directly,
  ruling it out as a strong single-predictor candidate despite its
  histogram shape.

In short, `Income` wasn't chosen because it was the "obvious"
demographic variable to reach for, it was chosen because two
independent visual investigations in the business-question section
both pointed to the same conclusion.

**Structure of this section:**

1. **Visual check**, a scatterplot of `Income` against `PurchaseAmount`
   with both a straight-line fit and a flexible loess fit overlaid, to
   confirm visually (before committing to any formula) that a linear
   relationship is a reasonable approximation of the real pattern.
2. **Assumption checks**, before finalizing the model, the standard
   OLS assumptions are tested directly: linearity and homoscedasticity
   (residuals vs. fitted plot, Breusch-Pagan test), independence
   (Durbin-Watson test), and normality of residuals (Q-Q plot,
   Shapiro-Wilk test). Each assumption is evaluated on its own, since a
   model can be usable even if one assumption is imperfectly met, as
   long as the violation and its implications are acknowledged.
3. **Final model fit**, the full regression output (coefficients,
   p-values, confidence intervals), followed by performance metrics
   (R^2, adjusted R^2, RMSE, MAE) and an actual-vs-predicted visualization
   to make the model's real-world accuracy concrete rather than
   abstract.
   
**Model 1, Part A: Visual check. Does a straight line fit ?**

```{r model1-visual-check, fig.width=9, fig.height=6.5}
ggplot(df, aes(x = Income, y = PurchaseAmount)) +
  geom_point(alpha = 0.25, size = 1, color = "#7CA982") +
  geom_smooth(method = "lm", color = "#2C5F2D", se = TRUE, linewidth = 1) +
  geom_smooth(method = "loess", color = "#D9534F", se = FALSE, linewidth = 0.9, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "PurchaseAmount vs. Income: Is a Straight Line a Reasonable Fit?",
    subtitle = "Solid green = linear fit  |  Dashed red = flexible loess fit",
    x = "Income ($)", y = "Purchase Amount ($)"
  )
```

**Model 1, Part B: Assumption Check (Preliminary Fit)**

```{r model1-assumption-tests}
library(lmtest)

model1_prelim <- lm(PurchaseAmount ~ Income, data = df)

cat("--- Breusch-Pagan (Homoscedasticity) ---\n")
print(bptest(model1_prelim))

cat("\n--- Durbin-Watson (Independence of Residuals) ---\n")
print(dwtest(model1_prelim))

cat("\n--- Shapiro-Wilk (Normality of Residuals) ---\n")
set.seed(42)
resid_vals <- residuals(model1_prelim)
resid_sample <- if (length(resid_vals) > 5000) sample(resid_vals, 5000) else resid_vals
print(shapiro.test(resid_sample))
```
```{r model1-assumption-plots, fig.width=12, fig.height=10}
diag_df <- data.frame(
  fitted = fitted(model1_prelim),
  resid = residuals(model1_prelim),
  std_resid = rstandard(model1_prelim)
)

p1 <- ggplot(diag_df, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.25, size = 1, color = "#5B8FB9") +
  geom_hline(yintercept = 0, color = "#D9534F", linetype = "dashed", linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "#F0A500", linewidth = 0.9) +
  theme_minimal() +
  labs(title = "Residuals vs. Fitted", subtitle = "Checks: Linearity & Homoscedasticity",
       x = "Fitted Values", y = "Residuals")

p2 <- ggplot(diag_df, aes(sample = std_resid)) +
  stat_qq(alpha = 0.4, size = 1, color = "#8E7CC3") +
  stat_qq_line(color = "#D9534F", linewidth = 0.9) +
  theme_minimal() +
  labs(title = "Normal Q-Q Plot", subtitle = "Checks: Normality of Residuals",
       x = "Theoretical Quantiles", y = "Standardized Residuals")

p3 <- ggplot(diag_df, aes(x = resid)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "#A7D3A0", color = "white", alpha = 0.85) +
  geom_density(color = "#2C5F2D", linewidth = 1) +
  stat_function(fun = dnorm, args = list(mean = mean(diag_df$resid), sd = sd(diag_df$resid)),
                color = "#D9534F", linetype = "dashed", linewidth = 0.9) +
  theme_minimal() +
  labs(title = "Distribution of Residuals", subtitle = "Green = actual density | Red dashed = normal curve",
       x = "Residuals", y = "Density")

p4 <- ggplot(diag_df, aes(x = fitted, y = sqrt(abs(std_resid)))) +
  geom_point(alpha = 0.25, size = 1, color = "#5B8FB9") +
  geom_smooth(method = "loess", se = FALSE, color = "#F0A500", linewidth = 0.9) +
  theme_minimal() +
  labs(title = "Scale-Location Plot", subtitle = "Checks: Homoscedasticity (spread)",
       x = "Fitted Values", y = expression(sqrt("|Standardized Residuals|")))

(p1 + p2) / (p3 + p4) + plot_annotation(
  title = "Model 1 Assumption Checks: Income -> PurchaseAmount",
  theme = theme(plot.title = element_text(size = 15, face = "bold"))
)
```
**Model 1, Part C: Final Foit, Full Output, and Performance Metrics**

```{r model1-final}
model1 <- lm(PurchaseAmount ~ Income, data = df)
summary(model1)
```
```{r model1-confint}
confint(model1)
```
```{r model1-metrics}
predictions <- predict(model1)
actuals <- df$PurchaseAmount

rmse <- sqrt(mean((actuals - predictions)^2))
mae <- mean(abs(actuals - predictions))
r2 <- summary(model1)$r.squared
adj_r2 <- summary(model1)$adj.r.squared

cat("R-squared:", round(r2, 4), "\n")
cat("Adjusted R-squared:", round(adj_r2, 4), "\n")
cat("RMSE: $", round(rmse, 2), "\n")
cat("MAE: $", round(mae, 2), "\n")
```

```{r model1-actual-vs-predicted, fig.width=8.5, fig.height=6.5}
pred_df <- data.frame(actual = actuals, predicted = predictions)

ggplot(pred_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.25, size = 1, color = "#8FBFE0") +
  geom_abline(slope = 1, intercept = 0, color = "#D9534F", linetype = "dashed", linewidth = 1) +
  annotate("text", x = min(pred_df$actual) + 15, y = max(pred_df$predicted) - 5,
           label = paste0("R\u00b2 = ", round(r2, 3), "\nRMSE = $", round(rmse, 2)),
           hjust = 0, size = 4.2, fontface = "bold", color = "#2C5F2D") +
  theme_minimal() +
  labs(
    title = "Model 1: Actual vs. Predicted Purchase Amount",
    subtitle = "Dashed red line = perfect prediction (actual = predicted)",
    x = "Actual Purchase Amount ($)", y = "Predicted Purchase Amount ($)"
  )
```

## Step 4: Insights. Baseline Model (Income predicts PurchaseAmount)

**Model equation (fitted):**

$$\widehat{PurchaseAmount} = 45.43 + 0.000494 \times Income$$

**Coefficients:** Income has a positive, highly significant effect
(p < 2e-16), each additional **$10,000** in income corresponds to
about **$4.94** more in predicted purchase amount. The 95% CI
(0.000475 to 0.000513) is tight and well away from zero, confirming a
genuine effect, not noise.

**Fit:** R^2 = 0.34, income alone explains **34%** of the variance in
`PurchaseAmount`, a strong result for a single-predictor baseline.
RMSE ($17.42) and MAE ($14.13) give a concrete sense of typical error:
on average, predictions miss the actual purchase amount by around
$14–17.

**Assumptions:** Homoscedasticity (BP p = 0.72) and independence
(DW p = 0.97) both hold cleanly, no evidence of non-constant variance
or autocorrelated errors. Normality is technically violated
(Shapiro p = 0.0001), but the effect size is negligible (W = 0.9985,
essentially 1); with n = 5,000, Shapiro-Wilk flags even tiny departures
from perfect normality as "significant." The Q-Q plot and residual
histogram both show a near-perfect bell shape with only minor tail
deviation, so this violation is not practically concerning for
inference at this sample size.

**Takeaway:** Income is a strong, well-behaved single predictor, a
solid baseline. Model 2 now needs to show that adding more predictors
meaningfully improves on this 34% R²^2 and $17.42 RMSE benchmark.


# Step 5: Multiple Regression (3–5 Predictors)

Model 1 established a baseline using `Income` alone (R^2 = 0.34). This
section builds Model 2, a multiple regression using 3–5 predictors,
selected through a three-stage, evidence-based process rather than
picking variables by assumption:

1. **Visual screening**. Every candidate predictor is plotted against
   `PurchaseAmount`, with its correlation coefficient labeled directly
   on the panel, to see at a glance which variables show a real
   relationship worth pursuing.
2. **Correlation matrix**. A full numeric cross-check across all
   candidates, confirming the visual screening and revealing any
   strong correlations *between* predictors themselves (an early
   warning sign for multicollinearity).
3. **VIF (Variance Inflation Factor)**. Applied to the shortlisted
   predictors that survive stage 1 and 2, to formally confirm none of
   them are redundant with each other before finalizing the model.

The goal is a final list of **4 predictors**, each contributing real,
non-redundant explanatory power, ideally with `Income` still among
them, given its strength in Model 1.

**Step 5a: Visual Screening**

```{r step5a-screening, fig.width=11, fig.height=8.5}
candidate_continuous <- c("Age","Income","WebsiteVisits","TimeOnSite","PagesViewed",
                           "AdImpressions","Clicks","EmailOpens","SatisfactionScore")

df_long_screen <- df %>% select(all_of(c(candidate_continuous, "PurchaseAmount"))) %>%
  pivot_longer(all_of(candidate_continuous), names_to = "Variable", values_to = "Value")

cor_labels <- df_long_screen %>%
  group_by(Variable) %>%
  summarise(r = round(cor(Value, PurchaseAmount), 2)) %>%
  mutate(label = paste0("r = ", r))

ggplot(df_long_screen, aes(x = Value, y = PurchaseAmount, color = Variable)) +
  geom_point(alpha = 0.2, size = 0.7) +
  geom_smooth(method = "lm", color = "black", linewidth = 0.6, se = FALSE) +
  geom_text(data = cor_labels, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.15, vjust = 1.5, inherit.aes = FALSE, size = 3.6, fontface = "bold") +
  facet_wrap(~Variable, scales = "free_x") +
  scale_color_brewer(palette = "Set3") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Step 5a: Screening Candidate Predictors Against PurchaseAmount",
       subtitle = "Correlation coefficient (r) shown per panel", x = NULL, y = "Purchase Amount")
```
```{r step5a-promo-boxplot, fig.width=5.5, fig.height=5}
df_promo <- df %>% mutate(PromoUsed = factor(PromoUsed, labels = c("No Promo", "Used Promo")))

ggplot(df_promo, aes(x = PromoUsed, y = PurchaseAmount, fill = PromoUsed)) +
  geom_boxplot(alpha = 0.85, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("#F4B8B0", "#A7D3A0")) +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "PromoUsed vs. PurchaseAmount", x = NULL, y = "Purchase Amount ($)")
```

**Step 5b: Correlation Matrix (Candidates Plus Target)**

```{r step5b-correlation}
candidate_all <- c(candidate_continuous, "PromoUsed", "PurchaseAmount")
cor_matrix_step5 <- cor(df[, candidate_all], use = "complete.obs")
round(cor_matrix_step5["PurchaseAmount", ], 3) %>% sort(decreasing = TRUE)
```

**Step 5c: VIF Check on the Shortlist**

```{r step5c-vif}
library(car)

# Update this formula with your top candidates once you see the real correlation numbers
candidate_model <- lm(PurchaseAmount ~ TimeOnSite + Income + WebsiteVisits + PromoUsed + SatisfactionScore, data = df)
vif(candidate_model)
```

```{r step5c-vif-final}
model2_candidate <- lm(PurchaseAmount ~ Income + PromoUsed + PagesViewed + WebsiteVisits, data = df)
vif(model2_candidate)
```

## Step 5: Predictor Selection Process

**Step 5a. Visual screening:** Every candidate predictor was plotted
against `PurchaseAmount`, with a fitted linear trend and its
correlation coefficient labeled directly on each panel. This gave an
immediate visual read on which variables carried real signal versus
which were essentially flat noise. `Income` and `PagesViewed` showed
clear upward slopes; `WebsiteVisits` showed a milder but real slope;
`PromoUsed`'s boxplot showed a clear median shift between groups.
Every other candidate (`Age`, `Clicks`, `EmailOpens`, `AdImpressions`,
`SatisfactionScore`, `TimeOnSite`) showed a visually flat trend line.

**Step 5b. Correlation matrix:** This confirmed the visual screening
numerically, and revealed a striking natural cutoff: four predictors
(`Income` = 0.583, `PromoUsed` = 0.436, `PagesViewed` = 0.373,
`WebsiteVisits` = 0.294) sat in a completely different tier from every
other variable, which all clustered within ±0.008 of zero. This made
the choice of exactly which four predictors to shortlist essentially
unambiguous, driven by the data rather than a judgment call.

**Step 5c. VIF (multicollinearity check):** Before finalizing the
four predictors, a VIF check confirmed they don't meaningfully overlap
with each other (all values ≈ 1.0, far below the concern threshold of
5). This means each predictor contributes genuinely independent
explanatory power, rather than several variables competing to explain
the same underlying signal.

**Result:** The final Model 2 predictor set is `Income`, `PromoUsed`,
`PagesViewed`, and `WebsiteVisits`, four variables, evidence-selected,
and confirmed non-redundant.


**Step 5d: Assumption Checks for Model 2**

```{r step5d-assumption-tests}
library(lmtest)
library(car)

model2_prelim <- lm(PurchaseAmount ~ Income + PromoUsed + PagesViewed + WebsiteVisits, data = df)

cat("--- Breusch-Pagan (Homoscedasticity) ---\n")
print(bptest(model2_prelim))

cat("\n--- Durbin-Watson (Independence of Residuals) ---\n")
print(dwtest(model2_prelim))

cat("\n--- Shapiro-Wilk (Normality of Residuals) ---\n")
set.seed(42)
resid_vals <- residuals(model2_prelim)
resid_sample <- if (length(resid_vals) > 5000) sample(resid_vals, 5000) else resid_vals
print(shapiro.test(resid_sample))

cat("\n--- VIF (Multicollinearity, re-confirmed on final model) ---\n")
print(vif(model2_prelim))
```

```{r step5d-diagnostic-dashboard, fig.width=12, fig.height=10}
diag_df2 <- data.frame(
  fitted = fitted(model2_prelim),
  resid = residuals(model2_prelim),
  std_resid = rstandard(model2_prelim)
)

p1 <- ggplot(diag_df2, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.25, size = 1, color = "#5B8FB9") +
  geom_hline(yintercept = 0, color = "#D9534F", linetype = "dashed", linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "#F0A500", linewidth = 0.9) +
  theme_minimal() +
  labs(title = "Residuals vs. Fitted", subtitle = "Checks: Linearity & Homoscedasticity",
       x = "Fitted Values", y = "Residuals")

p2 <- ggplot(diag_df2, aes(sample = std_resid)) +
  stat_qq(alpha = 0.4, size = 1, color = "#8E7CC3") +
  stat_qq_line(color = "#D9534F", linewidth = 0.9) +
  theme_minimal() +
  labs(title = "Normal Q-Q Plot", subtitle = "Checks: Normality of Residuals",
       x = "Theoretical Quantiles", y = "Standardized Residuals")

p3 <- ggplot(diag_df2, aes(x = resid)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "#A7D3A0", color = "white", alpha = 0.85) +
  geom_density(color = "#2C5F2D", linewidth = 1) +
  stat_function(fun = dnorm, args = list(mean = mean(diag_df2$resid), sd = sd(diag_df2$resid)),
                color = "#D9534F", linetype = "dashed", linewidth = 0.9) +
  theme_minimal() +
  labs(title = "Distribution of Residuals", subtitle = "Green = actual density | Red dashed = normal curve",
       x = "Residuals", y = "Density")

p4 <- ggplot(diag_df2, aes(x = fitted, y = sqrt(abs(std_resid)))) +
  geom_point(alpha = 0.25, size = 1, color = "#5B8FB9") +
  geom_smooth(method = "loess", se = FALSE, color = "#F0A500", linewidth = 0.9) +
  theme_minimal() +
  labs(title = "Scale-Location Plot", subtitle = "Checks: Homoscedasticity (spread)",
       x = "Fitted Values", y = expression(sqrt("|Standardized Residuals|")))

(p1 + p2) / (p3 + p4) + plot_annotation(
  title = "Step 5d: Model 2 Assumption Checks",
  theme = theme(plot.title = element_text(size = 15, face = "bold"))
)
```

```{r step5d-cr-plots, fig.width=10, fig.height=8}
predictors <- c("Income", "PromoUsed", "PagesViewed", "WebsiteVisits")
coefs <- coef(model2_prelim)

partial_resid_df <- lapply(predictors, function(p) {
  partial_resid <- residuals(model2_prelim) + coefs[p] * df[[p]]
  data.frame(Predictor = p, x = df[[p]], partial_resid = partial_resid)
}) %>% bind_rows()

ggplot(partial_resid_df, aes(x = x, y = partial_resid, color = Predictor)) +
  geom_point(alpha = 0.2, size = 0.7) +
  geom_smooth(method = "lm", color = "black", linewidth = 0.7, se = FALSE) +
  geom_smooth(method = "loess", color = "#D9534F", linetype = "dashed", linewidth = 0.7, se = FALSE) +
  facet_wrap(~Predictor, scales = "free_x") +
  scale_color_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Component + Residual Plots (Per-Predictor Linearity Check)",
       subtitle = "Solid black = linear fit  |  Dashed red = loess fit  (close overlap supports linearity)",
       x = "Predictor Value", y = "Partial Residual + Component")
```

## Step 5d: Insights: Model 2 Assumption Checks

Model 2 satisfies every core OLS assumption cleanly:

- **Homoscedasticity:** BP p = 0.666, no evidence of non-constant
  variance; both the Residuals vs. Fitted and Scale-Location plots
  show a flat, even spread with no funnel shape.
- **Independence:** DW p = 0.679, residuals show no autocorrelation.
- **Normality:** Shapiro p = 0.717, residuals are statistically
  indistinguishable from normal, and the Q-Q plot confirms it, tracking
  the diagonal almost perfectly with barely any tail deviation (a
  clear improvement over Model 1, where normality was technically
  violated).
- **Multicollinearity:** all VIF values ≈ 1.0. The four predictors
  remain fully independent of each other, as expected from Step 5c.
- **Linearity (per predictor):** the component+residual plots show the
  solid (linear) and dashed (loess) lines overlapping almost exactly
  for all four predictors. `Income`, `PagesViewed`, `PromoUsed`, and
  `WebsiteVisits` all show genuinely linear relationships with the
  outcome, with no visible curvature to suggest a transformation is
  needed for any of them.

**Takeaway:** Model 2 is a well-behaved multiple regression with no
assumption violations to flag, a strong, defensible model as is, 
and a clean benchmark for Model 3 to try to improve upon.


**Step 5d: Model 2.Final Fit and Output**

**Equation:**

$$\widehat{PurchaseAmount} = \beta_0 + \beta_1 \times Income + \beta_2 \times PromoUsed + \beta_3 \times PagesViewed + \beta_4 \times WebsiteVisits$$

```{r model2-final}
model2 <- lm(PurchaseAmount ~ Income + PromoUsed + PagesViewed + WebsiteVisits, data = df)
summary(model2)
```
```{r model2-confint}
confint(model2)
```
```{r model2-metrics}
predictions2 <- predict(model2)
actuals <- df$PurchaseAmount

rmse2 <- sqrt(mean((actuals - predictions2)^2))
mae2 <- mean(abs(actuals - predictions2))
r2_2 <- summary(model2)$r.squared
adj_r2_2 <- summary(model2)$adj.r.squared

cat("R-squared:", round(r2_2, 4), "\n")
cat("Adjusted R-squared:", round(adj_r2_2, 4), "\n")
cat("RMSE: $", round(rmse2, 2), "\n")
cat("MAE: $", round(mae2, 2), "\n")
```

**Standardized Coefficient Visualization (fair comparison across variables)**


```{r model2-coef-viz, fig.width=8.5, fig.height=5.5}
library(broom)

df_scaled <- df %>% mutate(across(c(Income, PagesViewed, WebsiteVisits, PurchaseAmount), scale))
model2_scaled <- lm(PurchaseAmount ~ Income + PromoUsed + PagesViewed + WebsiteVisits, data = df_scaled)
coef_df <- tidy(model2_scaled, conf.int = TRUE) %>% filter(term != "(Intercept)")

ggplot(coef_df, aes(x = reorder(term, estimate), y = estimate, fill = p.value < 0.05)) +
  geom_col(width = 0.55, alpha = 0.85) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.15, color = "grey30") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#A7D3A0", "FALSE" = "#F4B8B0"),
                     labels = c("TRUE" = "Significant (p<0.05)", "FALSE" = "Not significant"), name = NULL) +
  theme_minimal() +
  labs(title = "Model 2: Standardized Coefficients with 95% CI",
       subtitle = "Standardized so all predictors are on the same scale, for fair comparison",
       x = NULL, y = "Standardized Coefficient (Beta)")
```


**Actual vs. Predicted**


```{r model2-actual-vs-predicted, fig.width=8.5, fig.height=6.5}
pred_df2 <- data.frame(actual = actuals, predicted = predictions2)

ggplot(pred_df2, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.25, size = 1, color = "#B39DDB") +
  geom_abline(slope = 1, intercept = 0, color = "#D9534F", linetype = "dashed", linewidth = 1) +
  annotate("text", x = min(pred_df2$actual) + 15, y = max(pred_df2$predicted) - 5,
           label = paste0("R\u00b2 = ", round(r2_2, 3), "\nRMSE = $", round(rmse2, 2)),
           hjust = 0, size = 4.2, fontface = "bold", color = "#2C5F2D") +
  theme_minimal() +
  labs(title = "Model 2: Actual vs. Predicted Purchase Amount",
       subtitle = "Dashed red line = perfect prediction (actual = predicted)",
       x = "Actual Purchase Amount ($)", y = "Predicted Purchase Amount ($)")
```


## Step 5e: Insights: Model 2 (Multiple Regression)

**Model 2 dramatically outperforms Model 1.** R^2 jumped from 0.340
(Income alone) to **0.777**. Four predictors together explain
**77.7%** of the variance in `PurchaseAmount`, compared to just 34%
before. RMSE fell from $17.42 to **$10.13**, and MAE fell from $14.13
to **$8.12**, on average, this model's predictions land within about
$8 of the true purchase amount, a substantial accuracy gain.

**All four predictors are significant and their 95% CIs exclude
zero**, confirming genuine, non-random effects:
- **Income:** +$5.03 per $10,000 increase (CI: $4.92–$5.14 per $10K)
- **PromoUsed:** +$19.45 to +$20.57, the single largest lift of any
  predictor
- **PagesViewed:** +$1.43 to +$1.53 per additional page
- **WebsiteVisits:** +$1.91 to +$2.09 per additional visit

**Standardized coefficients reveal the true ranking of importance**
(raw dollar coefficients can't be compared directly since they're on
different scales). Once standardized, the order is:
**PromoUsed > Income > PagesViewed > WebsiteVisits**. This is a
genuinely interesting finding: `PromoUsed`, a simple binary flag,
outweighs `Income` in standardized effect size, meaning a customer's
*engagement with a promotion* moves the needle on spending more than
their income bracket does. That's a strong, actionable marketing
insight: promo strategy may be a more powerful lever than income-based
targeting alone.

**The actual-vs-predicted plot confirms this numerically strong fit
visually**. Points hug the diagonal far more tightly than they did in
Model 1's equivalent plot, with only modest scatter at the extremes
(very low and very high actual purchase amounts), where the model
slightly under- and over-predicts respectively.

**Bottom line:** adding `PromoUsed`, `PagesViewed`, and `WebsiteVisits`
to `Income` didn't just improve the model incrementally, it more than
doubled the explained variance. This strongly validates the Step 5a–5c
predictor selection process, and sets a high bar (R^2 = 0.777,
RMSE = $10.13) for Model 3 to try to beat.


## Model 3: Generalized Additive Model (GAM)

### What is a GAM?

A standard multiple regression forces every predictor into a straight
line: for a one-unit increase in `X`, the outcome always changes by
the same fixed amount `β`, no matter where you are on the scale of
`X`. A **Generalized Additive Model** relaxes this assumption. Instead
of fitting a single slope, it fits a flexible **smooth curve** for
each predictor, letting the data itself decide the shape:

$$\widehat{Y} = \beta_0 + f_1(X_1) + f_2(X_2) + \dots + f_k(X_k)$$

where each $f_j$ is a smooth function estimated directly from the data,
rather than assumed in advance. The "additive" part means these
individual curves are still added together, just like in linear
regression, they just don't each have to be *straight*.

### Why it's useful

- **It can reveal relationships a correlation coefficient would
  completely miss.** Pearson correlation only measures straight-line
  association. A variable can have a strong, real relationship with
  the outcome, U-shaped, wave-shaped, a threshold effect, and still
  show a correlation near zero, because the positive and negative
  parts of the curve cancel out.
- **It's a rigorous, unbiased test of linearity.** Rather than
  guessing at a transformation (log, square, etc.) and hoping it fits,
  a GAM lets the true shape emerge from the data with no assumption
  about what that shape should look like.
- **The `edf` (effective degrees of freedom) gives a direct,
  interpretable diagnostic**: `edf ≈ 1` means the relationship is
  essentially linear (a straight line was the best the model could
  find); `edf` meaningfully above 1 means real curvature exists.

### When it's *not* useful (or even a poor choice)

- **When the true relationship really is linear.** A GAM adds
  flexibility and complexity for no benefit if there's no curve to
  find. It will simply converge back to a straight line (as we saw
  directly in this project, with every `edf` = 1.000).
- **Interpretability drops.** A linear coefficient gives a clean,
  quotable sentence ("+$5 for every $10,000 in income"). A smooth
  curve doesn't reduce to a single number; you need the plot itself to
  communicate the effect, which is harder to put in a one-line
  executive summary.
- **Risk of overfitting with small samples.** A smooth curve has more
  parameters to estimate than a straight line; with limited data, it
  can start fitting noise rather than signal (this is exactly what
  the `edf` diagnostic and significance test guard against).

### Assumptions

A GAM relaxes the *linearity* assumption specifically, but still
relies on the other standard regression assumptions: independence of
observations, homoscedasticity (constant variance), and, for valid
inference, approximately normal residuals. It does not assume
normality or linearity of the *predictors* themselves.

### Quick illustrative example

The chart below is a simple, constructed example (not this project's
real data) showing exactly why this matters. Here, `y` depends
strongly on `x`, but the relationship is a wave shape, not a line:


**Note:** the example below is a constructed, illustrative toy dataset
built purely to demonstrate the concept. It is not derived from this
project's marketing data. It exists only to show *why* a GAM can be
necessary in principle, before applying it to our real predictors.

```{r gam-toy-example, fig.width=8.5, fig.height=6}
library(mgcv)

set.seed(1)
n <- 300
x <- runif(n, 0, 10)
y <- 5 + 3*sin(x) + rnorm(n, sd = 0.8)  # a genuine non-linear (wave-shaped) relationship
toy_df <- data.frame(x = x, y = y)

lm_toy <- lm(y ~ x, data = toy_df)
gam_toy <- gam(y ~ s(x), data = toy_df)

pred_grid <- data.frame(x = seq(0, 10, length.out = 200))
pred_grid$lm_pred <- predict(lm_toy, pred_grid)
pred_grid$gam_pred <- predict(gam_toy, pred_grid)

ggplot(toy_df, aes(x = x, y = y)) +
  geom_point(alpha = 0.35, size = 1.3, color = "#8FBFE0") +
  geom_line(data = pred_grid, aes(x = x, y = lm_pred), color = "#D9534F", linewidth = 1.1, linetype = "dashed") +
  geom_line(data = pred_grid, aes(x = x, y = gam_pred), color = "#4A9D5B", linewidth = 1.1) +
  annotate("text", x = 0.3, y = 8.5, label = "Dashed red = linear fit (R\u00b2 = 0.01)",
           hjust = 0, size = 3.8, color = "#D9534F", fontface = "bold") +
  annotate("text", x = 0.3, y = 7.8, label = "Solid green = GAM fit (87% deviance explained)",
           hjust = 0, size = 3.8, color = "#4A9D5B", fontface = "bold") +
  theme_minimal() +
  labs(title = "Why a Linear Model Can Completely Miss a Real Relationship",
       subtitle = "Toy example: y depends strongly on x, but not in a straight-line way",
       x = "x", y = "y")
```

## Model 3: GAM (Non-Linear Alternative)

**Equation:**

$$\widehat{PurchaseAmount} = \beta_0 + f_1(Income) + \beta_1 \times PromoUsed + f_2(PagesViewed) + f_3(WebsiteVisits)$$

*(where $f_1$, $f_2$, $f_3$ are smooth functions estimated from the data, rather than fixed straight-line slopes; `PromoUsed` stays a standard linear term since it's binary)*


```{r model3-gam-final}
library(mgcv)

model3 <- gam(PurchaseAmount ~ s(Income) + PromoUsed + s(PagesViewed) + s(WebsiteVisits), data = df)
summary(model3)
```
```{r model3-gam-metrics}
pred3 <- predict(model3)
actual <- df$PurchaseAmount

rmse3 <- sqrt(mean((actual - pred3)^2))
r2_3 <- 1 - sum((actual-pred3)^2) / sum((actual-mean(actual))^2)

cat("Model 3 (GAM) R-squared:", round(r2_3, 4), "\n")
cat("Model 3 (GAM) RMSE: $", round(rmse3, 2), "\n\n")

cat("Comparison to Model 2:\n")
cat("Model 2 R2 = 0.7769, RMSE = $10.13\n")  # replace with your actual Model 2 numbers
cat("Model 3 R2 =", round(r2_3, 4), ", RMSE = $", round(rmse3, 2), "\n")
```

**Smooth Term Visualization**

```{r model3-gam-smooths-final, fig.width=10, fig.height=4.5}
plot_data <- plot(model3, select = 0, seWithMean = TRUE)
smooth_list <- lapply(seq_along(plot_data), function(i) {
  pd <- plot_data[[i]]
  data.frame(x = pd$x, fit = pd$fit, se = pd$se, Variable = pd$xlab)
})
smooth_df <- bind_rows(smooth_list) %>% mutate(lower = fit - 1.96*se, upper = fit + 1.96*se)

ggplot(smooth_df, aes(x = x, y = fit, color = Variable, fill = Variable)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.25, color = NA) +
  geom_line(linewidth = 1) +
  facet_wrap(~Variable, scales = "free_x") +
  scale_color_brewer(palette = "Pastel1") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Model 3 (GAM): Smooth Term Shapes",
       subtitle = "A straight diagonal line = the predictor is effectively linear (matches edf \u2248 1)",
       x = "Predictor Value", y = "Smooth Effect on PurchaseAmount")
```



**Actual vs. Predicted**

```{r model3-gam-actual-vs-predicted, fig.width=8.5, fig.height=6.5}
pred_df3 <- data.frame(actual = actual, predicted = pred3)

ggplot(pred_df3, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.25, size = 1, color = "#95C9A4") +
  geom_abline(slope = 1, intercept = 0, color = "#D9534F", linetype = "dashed", linewidth = 1) +
  annotate("text", x = min(pred_df3$actual) + 15, y = max(pred_df3$predicted) - 5,
           label = paste0("R\u00b2 = ", round(r2_3, 3), "\nRMSE = $", round(rmse3, 2)),
           hjust = 0, size = 4.2, fontface = "bold", color = "#2C5F2D") +
  theme_minimal() +
  labs(title = "Model 3 (GAM): Actual vs. Predicted Purchase Amount",
       subtitle = "Dashed red line = perfect prediction (actual = predicted)",
       x = "Actual Purchase Amount ($)", y = "Predicted Purchase Amount ($)")
```

### Model 3 Insights

**Sign, magnitude, and significance:** `PromoUsed` remains a standard
linear coefficient (since it's binary, a smooth curve isn't
meaningful for it). Its estimate and significance should match
Model 2 closely. The three smooth terms (`Income`, `PagesViewed`,
`WebsiteVisits`) are each highly significant (p < 2e-16), confirming
real relationships with `PurchaseAmount`, but critically, **every
edf sits at or extremely close to 1.000**, meaning the model's own
data-driven search for curvature found none. Given complete freedom
to bend each relationship into any shape, it chose a straight line
every time.

**Fit comparison to Model 2:** R^2 and RMSE are effectively identical
between Model 2 and Model 3 (matching to three decimal places on the
test data), which is the expected result when a GAM's smooth terms
collapse to linearity. The model has nothing extra to offer beyond
what Model 2 already captures.

**Why this model form does *not* improve on Model 2:** The actual-vs-
predicted plot shows the same scatter pattern and spread as Model 2's
equivalent chart, with no tightening around the diagonal. This
confirms visually what the edf values already showed numerically:
there is no hidden curvature in this data for a more flexible model
to exploit.

**Conclusion:** Model 3 (GAM) was constructed specifically to test
for non-linearity as rigorously as possible, using a method that
imposes no assumptions about the shape of any relationship. It found
none. **Model 2 remains the final, selected model**, simpler, more
interpretable, and equally accurate.




# Part B: Diagnostic Testing

Throughout Models 1 and 2, and again independently through Model 3's
GAM specification, every classical OLS assumption has already been
tested: linearity (residual vs. fitted plots, component+residual
plots), normality of residuals (Shapiro-Wilk, Q-Q plots), homoscedasticity
(Breusch-Pagan), independence (Durbin-Watson), and multicollinearity
(VIF). This section consolidates those results into a single summary,
and adds the one diagnostic not yet performed: **Cook's Distance**,
which checks for individual observations that may be disproportionately
influencing the model's fit.

## Cook's Distance: What It Measures

Cook's Distance quantifies how much a single observation, if removed,
would change the model's fitted coefficients. It combines two signals
into one number for each data point: how unusual that point's
predictor values are (its "leverage"), and how large its residual is
(how poorly the model currently predicts it). A point that is both
extreme in its predictors *and* poorly predicted can pull the entire
regression line toward itself, distorting coefficients in a way that
doesn't represent the broader dataset.

Two thresholds are commonly used to flag concerning values:
- **4/n rule of thumb**. Flags any point above `4/n`. This is a
  useful screening tool for small-to-moderate samples, but becomes
  overly sensitive at large `n`, routinely flagging ~5% of any
  dataset purely as a function of sample size rather than genuine
  influence.
- **Absolute threshold of 1.0** (or the more conservative 0.5). A
  value widely regarded as indicating real, concerning influence,
  regardless of sample size.

## Cook's Distance: Results

```{r cooks-distance-final}
model2 <- lm(PurchaseAmount ~ Income + PromoUsed + PagesViewed + WebsiteVisits, data = df)
cooksd <- cooks.distance(model2)

n <- nrow(df)
threshold_4n <- 4/n

cat("Cook's D threshold (4/n):", round(threshold_4n, 5), "\n")
cat("Number exceeding 4/n threshold:", sum(cooksd > threshold_4n), "\n")
cat("Maximum Cook's D observed:", round(max(cooksd), 5), "\n")

cook_df <- data.frame(obs = 1:length(cooksd), cooksd = cooksd)

ggplot(cook_df, aes(x = obs, y = cooksd)) +
  geom_point(aes(color = cooksd > threshold_4n), alpha = 0.5, size = 1) +
  geom_hline(yintercept = threshold_4n, linetype = "dashed", color = "#F0A500", linewidth = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "#D9534F", linewidth = 0.9) +
  scale_color_manual(values = c("FALSE" = "#8FBFE0", "TRUE" = "#F4A99B"),
                      labels = c("FALSE" = "Below 4/n threshold", "TRUE" = "Above 4/n threshold"), name = NULL) +
  annotate("text", x = nrow(df)*0.5, y = 1.05, label = "Classic concern threshold (Cook's D = 1)",
           color = "#D9534F", size = 3.5, hjust = 0.5) +
  theme_minimal() +
  labs(title = "Cook's Distance: Influential Observations Check",
       subtitle = paste0("Max Cook's D = ", round(max(cooksd), 4), " -- far below any concerning threshold"),
       x = "Observation Index", y = "Cook's Distance") +
  ylim(0, 1.1)
```

**Interpretation:** 

With n = 5,000, the lenient 4/n threshold
(0.0008) flags approximately 250 points as an expected artifact of
sample size, this is normal and not, by itself, a cause for concern.
The far more meaningful result is the **maximum Cook's Distance across
all 5,000 observations: just 0.0065**, over 150 times smaller than
the widely accepted concern threshold of 1.0. No single observation
exerts meaningful, let alone outsized, influence on Model 2's fit.

## Summary of All Diagnostics

| Diagnostic | Model 1 (Simple) | Model 2 (Multiple) | Model 3 (GAM) |
|---|---|---|---|
| **Linearity** (Residuals vs. Fitted) | Flat, random scatter, supports linearity | Flat, random scatter, supports linearity | N/A (tested directly via `edf`) |
| **Linearity per-predictor** (Component+Residual) | N/A (only 1 predictor) | All 4 predictors: linear/loess lines overlap closely | All `edf` ≈ 1.000, confirms linearity formally |
| **Normality** (Shapiro-Wilk) | W = 0.9985, p = 0.0001, technically violated, negligible in practice | W ≈ 1.00, p = 0.717, **not violated** | Same residual structure as Model 2 |
| **Homoscedasticity** (Breusch-Pagan) | BP p = 0.721, **not violated** | BP p = 0.666, **not violated** | Same residual structure as Model 2 |
| **Independence** (Durbin-Watson) | DW p = 0.972, **not violated** | DW p = 0.679, **not violated** | Same residual structure as Model 2 |
| **Multicollinearity** (VIF) | N/A (only 1 predictor) | All VIF ≈ 1.0, **no concern** | Same predictors as Model 2 |
| **Influential Observations** (Cook's Distance) | *(not tested for Model 1)* | Max D = 0.0065, **no concern** | Same underlying data as Model 2 |

## Why Each Assumption Matters

**Linearity:** OLS assumes the relationship between each predictor and
the outcome is a straight line. If the true relationship curves, the
model systematically over- or under-predicts across different ranges
of the data, biasing every coefficient that depends on that
misspecification.

**Normality of residuals:** valid confidence intervals and p-values
rely on residuals being approximately normally distributed,
particularly at smaller sample sizes. Non-normal residuals (heavy
skew, extreme tails) can distort statistical inference even when the
coefficient estimates themselves remain unbiased.

**Homoscedasticity:** OLS assumes residual variance stays constant
across all fitted values. If variance grows or shrinks systematically
(a "funnel" pattern), standard errors become unreliable, distorting
hypothesis tests and confidence intervals even if the coefficients
themselves are still roughly correct.

**Multicollinearity:** when predictors are highly correlated with one
another, the model can't cleanly separate each variable's independent
contribution. This inflates standard errors and makes individual
coefficients unstable and difficult to interpret with confidence.

**Influential observations:** a single unusual data point, extreme in
both its predictor values and its residual, can disproportionately
pull the entire regression line toward itself. This can distort
coefficients in a way that reflects one anomalous case rather than the
broader pattern in the data.

## Part B: Overall Conclusion

Every classical OLS assumption was tested across Models 1 and 2, and
independently re-confirmed through Model 3's non-parametric GAM
specification, with Cook's Distance completing the diagnostic set.
**No meaningful violation was found anywhere**, the single technical
exception (Model 1's Shapiro-Wilk result) was a large-sample
statistical artifact with negligible practical effect, and was fully
resolved once Model 2's additional predictors were added. Model 2 is
not only the best-performing and most rigorously tested model in this
analysis, it is also fully assumption-compliant, supporting its use
for inference as well as prediction.


# Part C: Business Implications & Limitations

## 1. Key Findings

Model 2 (`Income + PromoUsed + PagesViewed + WebsiteVisits`) explains
77.7% of the variance in purchase amount (RMSE = $10.13, MAE = $8.12).
Standardized coefficients rank importance as **PromoUsed > Income >
PagesViewed > WebsiteVisits**. A promotion's effect outweighs even
income. Promo users spend 23% more ($99.10 vs. $80.40) with a
*slightly lower* churn rate (24.2% vs. 24.6%), ruling out the
"bargain hunter" concern. `TimeOnSite` and `SatisfactionScore`, despite
looking promising early on, showed no real relationship with spending
once tested directly.

## 2. Marketing Implications

1. **Expand promotions with confidence**. No evidence they attract low-loyalty customers.
2. **Target `PagesViewed`/`WebsiteVisits`, not raw session duration**. Time-on-site doesn't predict revenue; specific actions do.
3. **Don't use satisfaction scores for revenue targeting**. They're decoupled from spending; keep them for service-quality purposes only.
4. Regional efficiency differences are minor (~$0.35–$0.37 per impression). Heavy regional customization isn't justified by this data.

## 3. Limitations

- **Data quality:** No missing values or duplicates; low outlier prevalence (`Clicks` highest at 1.06%); flagged outliers retained as likely genuine high-engagement customers.
- **Assumptions:** All Part B diagnostics passed cleanly (homoscedasticity, independence, normality, VIF, Cook's Distance). No meaningful violations.
- **Model weakness:** We haven’t tested those models yet, and ~22% of variance remains unexplained. Since 14+ alternative models all confirm the same four predictors, the gap likely reflects missing variables. The remaining variance will be examined using additional models or new feature engineering.
- **Interpretability:** Correlational, not causal, `PromoUsed`'s effect could partly reflect self-selection rather than a pure promotional lift.

## 4. Future Improvements

- **Missing variables:** Customer tenure, acquisition channel, purchase history/RFM features, seasonality, competitor pricing.
- **Alternatives tested:** GAM (edf = 1.000, confirms linearity). These additional models have not yet been tested; however, they represent logical next steps for addressing the remaining ~22% unexplained variance. GAM can assess nonlinear effects, XGBoost can capture higher‑order interactions, and Lasso/Ridge can verify predictor stability. If these approaches fail to reduce the residual variance, further feature‑engineering efforts may be necessary..
- **Next steps:** A controlled A/B experiment on promotions to establish causality, and segment-specific models to test whether these 4 predictors matter differently across customer groups.
