---
title: "Marketing Analytics Regression"
subtitle: "Regression Modeling of Customer Behavior"
author: "Jose Alberto Martinez Morales"
date: "2026-07-31"
output:
  html_document:
    toc: true
    toc_float: true
---

**Marketing Analytics (DAMO-520-21)**

**Instructor:** Prof. Mohamad Makki

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook applies regression modeling to `marketing_analytics_dataset_5000.csv`
to analyze customer behavior, following the assignment structure:

- **Part A**: Exploratory analysis and three regression models (baseline,
  multiple regression, improved/alternative form)
- **Part B**: Diagnostic testing (linearity, normality, homoscedasticity,
  multicollinearity, influential observations)
- **Part C**: Business implications and limitations (written up separately
  in the final report)

We'll build this up step by step, one chunk at a time.

# Step 0: Load Packages and Data

```{r load-packages}
library(tidyverse)   # dplyr, ggplot2, etc.
library(car)         # VIF, component+residual plots
library(lmtest)      # Breusch-Pagan test
library(GGally)      # correlation matrix / pairs plots
library(readxl)      # for reading .xlsx files
```

```{r load-data}
df <- read_excel("C:/Users/piuze/OneDrive/Documentos/marketing-analytics-regression/data/marketing_analytics_dataset_5000.xlsx")
dim(df)
str(df)
```

# Step 1: Exploratory Analysis

*(To fill in together \u2014 choosing the dependent variable, at least four
independent variables, correlation matrix, and distribution/relationship
plots.)*

# Step 2: Model Development

*(To fill in together \u2014 Model 1 baseline, Model 2 multiple regression,
Model 3 improved/alternative form.)*

# Part B: Diagnostic Testing

*(To fill in together \u2014 linearity, normality, homoscedasticity,
multicollinearity, influential observations.)*
